Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex
Some changes from Matt's comments on section 5

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

For this, we introduce a new value for a delayed lookup that acts to
quotient the state space by the values at the delayed address. Its
changes to the semantics are highlighted in the following rules:

\begin{align*}
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\}\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

\paragraph{Precision loss} This semantics introduces a subtle precision difference over the
baseline that is negligible. Consider a configuration where a
reference to a variable and a binding of a variable will happen in one
step. With laziness, the reference will mean the original binding(s)
of the variable or the new one, because the actual store lookup is
delayed precisely one step (i.e. laziness is administrative). Without
laziness, the reference will fan out to all the bindings of the
variable before the new binding happens and thus might have an
observable precision difference. In our experience, this situation
happens too rarely to warrant the significant increase in time and
memory needed for this eager non-determinism. Indeed, were the
variable reference a step later and another binding not made in
that step, the results of the two approaches are the same.

\paragraph{Avoiding precision loss} The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with store-allocated results consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

Laziness removes several represented states that would then have to be
stepped by the fixpoint combinator.  We can further eliminate the {\tt
  ev} states from the run-time interpretation of a program by
specializing the machine transition relation to the program being
analyzed. This precision-preserving transformation eliminates
interpretative overhead by first compiling the program into ``byte
code'' instructions.

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
    {\tt ev} states. The next step is to change creation of {\tt ev}
    states to simply calling these functions. The modified reduction
    relation is shown in figure \ref{fig:caam}.

\begin{figure}
\begin{align*}
\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\label{fig:compile}
\end{figure}

@@

\subsection{Store deltas}

Each step in the fixpoint for the above techniques requires joining
several, commonly large, stores together. Each step is unlikely to
always modify all addresses of the store, so this is needlessly
memory- and time-intensive. We can instead accumulate store changes
and apply them after all steps have completed. This uses far fewer
join operations, leading to much less allocation and computational
overhead.

We represent changes as $Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
begins as empty for each step. Applying the changes is
straightforward:

@@

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs\cup cs', \msdiff(\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

%% Delta Store + Compile + Lazy:
%%    cpu time: 668 real time: 686 gc time: 41

@@

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire
stores - an expensive computation.  We modify a technique due to
Shivers~\cite{ianjohnson:Shivers:1991:CFA}, timestamp approximation,
that removes the need for joining whole stores by manipulating a
global store imperatively. The original technique loses precision
because states are stepped in sequence, modifying the store in place -
more values or continuations can be added before another state gets to
step. This interference means more possible flows (i.e. less
precision) and a nondeterministic analysis - the order the set of states is
traversed is not necessarily deterministic. However, by delaying the
store changes until after all steps have completed, we can maintain
our precision and determinism. In a variety of experiments, we
observed that the time and space overhead for storing store-deltas was outweighed by
the precision keeping flows in check. That is to say, the analysis was
faster {\it and} more precise.

We pair this technique with a mutable workset to avoid the
intermediate sets introduced by the $State \times Store \to \Set(State) \times Store\msdiff$ signature
of $\machstep$. This necessitates a change to the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it would have otherwise accumulated in a set that would later be added to the workset.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

>>>>

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

For this, we introduce a new kind of value, $\saddr{\maddr}$, for delaying looking up $\maddr$. This acts to quotient the state space by the values at the delayed address. The following rules highlight the changes to the semantics:

\begin{align*}
force &: Store \times Addr \to \Set(Value) \\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\}\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

\paragraph{Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed precisely one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with store-allocated results consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

\paragraph{Our choice: option 1}
The configurations that lead to precision loss happen too rarely to
warrant the significant increase in time and memory needed for this
eager non-determinism. Indeed, were the variable reference a step
later and another binding not made in that step, the results of the
two approaches are the same.

@@

The laziness optimization saved time by doing the same amount of reasoning as before but in fewer transitions. We can exploit the same idea - same reasoning, fewer transitions - with abstract compilation. Abstract compilation transforms complex expression whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''  The abstract interpreter then does in one transition what previously took many. Abstract compilation eliminates unnecessary allocation, deallocation and branching.

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
    {\tt ev} states. The next step is to change creation of {\tt ev}
    states to simply calling these functions. Figure \ref{fig:caam}
    shows the modified reduction relation.

\begin{figure}
\begin{align*}
\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

@@

\subsection{Locally log-based store deltas}

Every step the analysis makes for the above techniques requires
joining several, commonly large, stores together. Each step is
unlikely to always modify all addresses of the store, so this is
needlessly memory- and time-intensive. We can instead log store
changes and replay the change log on the full store after all steps
have completed. This uses far fewer join operations, leading to less
overhead.

We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
begins empty ($\mtlst$) for each step. Applying the changes to the
full store is straightforward - we interpret $\msdiff$ as a $Store \to
Store$ function:

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', \msdiff(\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

Here $concat : \Set(Store\msdiff) \to Store\msdiff$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
concat(\varnothing) &= \mtlst \\
concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
\end{align*}

%% Delta Store +k Compile + Lazy:
%%    cpu time: 668 real time: 686 gc time: 41

@@

Thus far, we have made our optimizations in a purely functional
manner. For the next series of optimizations, we need to dip into the
imperative. We can motivate this entire sequence of optimizations by
focusing on the largest bottleneck in the current state-space
exploration algorithm: checking to see if a state has already been
seen. Given two states, checking equality is expensive because the
stores within each are large, and every entry must be checked against
every other. Hashes can sometimes rule out inequality relatively
quickly, but the incidence of collisions and actual equality is
costly.

And, there is a better way. Shivers' original work on k-CFA was
susceptible to the same problem, and he suggested three complementary
optimizations: (1) make the store global; (2) update the store
imperatively; and (3) associate every change in the store with a
version number -- its timestamp. Then, put timestamps in states
where previously there were stores. Given two states, the analysis can
now compare their stores just by comparing their timestamps -- a
constant-time operation.

There is a subtle loss of precision and Shivers' original time-stamp
technique that we can fix. For a given abstract state, all writes to
the global store need to be delayed until the analysis considers all
branches from that abstract state. This avoids cross-branch pollution
which would not otherwise happen, e.g., when one branch writes to
address $\maddr$ and another branch reads from address
$\maddr$. Fortunately, given our conversion to log-based stores this
change is straightforward.

The current state of the state-space search now has six components:

\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}
